\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2021-06-16 Articles/2020-QuantumLogic/Source/quantumlogic.tex
QLogic: minor edits from Gabriele

In this paper we will be concerned with a similar issue, since we ask whether standard quantum mechanics itself necessarily involves a new logical structure radically different with respect to classical propositional calculus. In what follows we will answer this question in the negative. More precisely, this essay has a two-fold aim: in the first place, we analyze the main motivations and justifications for the introduction of quantum logic, claiming that such arguments do not really entail the failure of CL in the quantum context because they contain serious flaws. In the second place, we show that from the mathematical structures on which QM is built, it is possible to provide a distributive lattice of quantum propositions, contrary to the received view in QL. Such a distributive lattice of quantum statements, moreover, allows us to take into account assertions concerning expectation values, which are generally excluded in the standard approach to quantum logic. Finally, we argue that it is possible to demonstrate that classical logical propositions can form a non-distributive lattice. All these aspects, to our knowledge, are not sufficiently discussed in literature, thus, the present paper can contribute to shed new light on the relation between logic and quantum mechanics.
\vspace{2mm}

The essay is structured as follows: in Section \ref{QL} we review the main arguments that have been advanced in literature to introduce a quantum logical calculus, here we also provide a brief outline of the standard approach to QL (readers familiar with Birkhoff and von Neumann's proposal can skip this part). Section \ref{Motivations} explains why the main examples used to show the failure of classical logic in QM contain important flaws and ambiguities; moreover, we argue that the classical distributive law can be maintained in quantum mechanics, once such ambiguities are removed. In Section \ref{Math}, contrary to the received view in QL, we show not only that quantum logical propositions can generate a distributive lattice, but also that important sentences of quantum theory---i.e.\ those concerning expectations values---are excluded by standard quantum logic.\ In Section \ref{conc} we discuss the main philosophical implications of the preceding sections and conclude the paper.

@@

This situation, however, is formally identical to the previous one: before we had that $q_o \vee r_o \neq q_i \vee r_i$, now we have $p_f \vee q_f \neq p_o \vee q_o$. It would be improper to say that an electron has always either $x$-spin up or $y$-spin up just because it is so after the third step. Similarly, it would be improper as well to claim that the electron has always either $y$-spin up or down just because it is so after the measurement of spin along the $y$ axis. Neither the disjunction, nor its atomic components tell us anything about an inherent property associated with the observable $S_y$ possessed by a quantum particle independently of any measurement of $S_y$.

@@

To sum up, \eqref{DL} fails in quantum domain because the we considered propositions evaluated at different times, before and after the measurement, in the two sides of the biconditional. If we remove the ambiguity and make the temporal dependence explicit, the distributivity law is recovered. The lack of commutativity of the observables in quantum mechanics therefore does not change the rules of logic. It imposes that the lattice of propositions at different times depends on the process, in particular on the choice of measurement. That is, if at time $t$ a particular observable is either prepared or measured, all statements about incompatible observables will be impossible and therefore they will be false. Quantum measurements do have an impact on the lattice of statements, unlike classical measurements, but that impact does not change the rules of logic.

@@

In order to begin our discussion, let us consider in the first place a set $X$: in the quantum case it will correspond to the set of all possible quantum states; similarly, in the classical case it will correspond to the set of all possible probability distributions over phase space.\footnote{We should stress, however, that $\sigma$-algebras are defined independently of what the elements of $X$ actually are. For a simpler example, $X$  could correspond to the set of the real numbers for a continuous quantity.} By definition, a $\sigma$-algebra is a collection of subsets of $X$ closed under countable union, countable intersection and complement with respect to $X$; the closure condition requires that  performing those operations on the sets contained in a $\sigma$-algebra will return another set in the same $\sigma$-algebra. Interestingly, these sets represent statements concerning the objects described by $X$, i.e.\ quantum states and classical probability distributions.\footnote{In the quantum case this is exemplified also in Section \ref{QL}.} Then, in virtue of the equivalence between sets and statements, it will be more convenient for us to conceive $\sigma$-algebras as collections of propositions about $X$ instead of sets. Referring to this, since the $\sigma$-algebra is closed under countable union, countable intersection and complement with respect to $X$, the corresponding statements will be closed under countable conjunction, countable disjunction and complement. Hence, the $\sigma$-algebra generates a Boolean algebra of statements.\footnote{Note that the interpretation used here is exactly the same used in probability theory. In fact, a probability space is defined by three elements: a sample space $\Omega$ which represents the set of all possible outcomes; a $\sigma$-algebra over $\Omega$ which represents all events, all propositions; and a measure that assigns a probability to each event.}

@@

While all of this may seem suspect to some reader, we would like to stress that these are very standard constructions. Alternatively stated, all of what has been mentioned so far are the standard mathematical techniques used to study these spaces and can be found in any textbook (cf.\ \cite{Rudin:1991}, \cite{Vasudeva:2017}). Here we are bringing out the full physical significance of these mathematical constructs, which is too often overlooked. Understanding the link between the mathematical structures and the physics they represent is crucial to clearing out possible misunderstandings.

>>>>

In this paper we will be concerned with a similar issue, since we ask whether standard quantum mechanics itself necessarily involves a new logical structure radically different with respect to classical propositional calculus. In what follows we will answer this question in the negative. More precisely, this essay has a two-fold aim: in the first place, we analyze the main motivations and justifications for the introduction of quantum logic, claiming that such arguments do not really entail the failure of CL in the quantum context because they contain serious flaws. In the second place, we show that the mathematical structure on which QM is built already provides us with a distributive lattice of quantum propositions, contrary to the received view in QL. Such a distributive lattice of quantum statements, moreover, allows us to take into account assertions concerning expectation values, which are generally excluded in the standard approach to quantum logic but are nonetheless experimentally relevant. Finally, we argue that classical probabilistic theory are also equipped with an analogue non-distributive lattice. All these aspects, to our knowledge, are not sufficiently discussed in literature, thus, the present paper can contribute to shed new light on the relation between logic and quantum mechanics.
\vspace{2mm}

The essay is structured as follows: in Section \ref{QL} we review the main arguments that have been advanced in literature to introduce a quantum logical calculus, here we also provide a brief outline of the standard approach to QL (readers familiar with Birkhoff and von Neumann's proposal can skip this part). Section \ref{Motivations} explains why the main examples used to show the failure of classical logic in QM contain important flaws and ambiguities; moreover, we argue that the classical distributive law can be maintained in quantum mechanics, once such ambiguities are removed. In Section \ref{Math}, contrary to the received view in QL, we show that quantum theory already posses a distributive lattice of propositions in the form of the Borel algebra, which includes not only the lattice of standard quantum logic, but also important sentences such as those concerning expectations values. We also show that classical probabilistic theory posses a formally analogue structure. In Section \ref{conc} we discuss the main philosophical implications of the preceding sections and conclude the paper.

@@

This situation is formally identical to the previous one: before we had that $q_o \vee r_o \neq q_i \vee r_i$, now we have $p_f \vee q_f \neq p_o \vee q_o$. It would be improper to say that an electron has always either $x$-spin up or $y$-spin up just because it is so after the third step. Similarly, it would be improper to claim that the electron has always either $y$-spin up or down just because it is so after the measurement of spin along the $y$ axis. Neither the disjunction, nor its atomic components tell us anything about an inherent property associated with the observable $S_y$ possessed by a quantum particle independently of any measurement of $S_y$.

@@

To sum up, \eqref{DL} fails in the quantum domain because we considered propositions evaluated at different times, before and after the measurement, in the two sides of the biconditional. If we remove the ambiguity and make the temporal dependence explicit, the distributivity law is recovered. The lack of commutativity of the observables in quantum mechanics therefore does not change the rules of logic. It imposes that the lattice of propositions at different times depends on the process, in particular on the choice of measurement. That is, if at time $t$ a particular observable is either prepared or measured, all statements about incompatible observables will be impossible and therefore they will be false. Quantum measurements do have an impact on the lattice of statements, unlike classical measurements, but that impact does not change the rules of logic.

@@

In order to begin our discussion, let us consider in the first place a set $X$: in the quantum case it will correspond to the set of all possible quantum states; similarly, in the classical case it will correspond to the set of all possible probability distributions over phase space.\footnote{We should stress, however, that $\sigma$-algebras are defined independently of what the elements of $X$ actually are. For a simpler example, $X$  could correspond to the set of the real numbers for a continuous quantity.} By definition, a $\sigma$-algebra is a collection of subsets of $X$ closed under countable union, countable intersection and complement with respect to $X$; the closure condition requires that  performing those operations on the sets contained in a $\sigma$-algebra will return another set in the same $\sigma$-algebra. Interestingly, these sets can be understood to represent statements concerning the objects described by $X$, i.e.\ quantum states and classical probability distributions.\footnote{In the quantum case this is exemplified also in Section \ref{QL}.} Namely, if $U \subseteq X$, the statement ``$x$ is in $U$'' tells us that the state $x$ must one of those delimited by $U$. Then, in virtue of the equivalence between sets and statements, it will be more convenient for us to conceive $\sigma$-algebras as collections of propositions about $X$ instead of sets. Referring to this, since the $\sigma$-algebra is closed under countable union, countable intersection and complement with respect to $X$, the corresponding statements will be closed under countable conjunction, countable disjunction and complement. Hence, the $\sigma$-algebra generates a Boolean algebra of statements.\footnote{Note that the interpretation used here is exactly the same used in probability theory. In fact, a probability space is defined by three elements: a sample space $\Omega$ which represents the set of all possible outcomes; a $\sigma$-algebra over $\Omega$ which represents all events, all propositions; and a measure that assigns a probability to each event.}

@@

While all of this may seem suspect to some readers, we would like to stress that these are very standard constructions. Alternatively stated, all of what has been mentioned so far are the standard mathematical techniques used to study these spaces and can be found in any textbook (cf.\ \cite{Rudin:1991}, \cite{Vasudeva:2017}). Here we are bringing out the full physical significance of these mathematical constructs, which is too often overlooked. Understanding the link between the mathematical structures and the physics they represent is crucial to clearing out possible misunderstandings.
